Exploring mathematical objects from custom-tailored mathematical universes

### 2018-10-30 paper-filmat.tex
Formal translation rules

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,stmaryrd}
\usepackage{booktabs}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}

@@

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}

@@

A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
Turing machine~$M$ which, given the description of such a Turing machine~$X$ as
input, outputs the description of a similar such Turing machine~$Y$ as output.
To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
and can therefore determine arbitrarily many rational approximations~$X(m)$.
However, within finite time, the machine~$Y$ can only acquire finitely many
such approximations. Hence a function such as the signum function, for which
even rough rational approximations of~$\sgn(x)$ require infinite precision in
the input~$x$, do not exist in the \effective topos.

% XXX formal rules

>>>>

\usepackage[utf8]{inputenc}
\usepackage{amsthm,mathtools,stmaryrd,amssymb}
\usepackage{booktabs}
\usepackage[all]{xy}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{xspace}

@@

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}

@@

A function~$f : \RR \to \RR$ in the \effective topos is therefore given by a
Turing machine~$M$ which, given the description of such a Turing machine~$X$ as
input, outputs the description of a similar such Turing machine~$Y$ as output.
To compute a rational approximation~$Y(n)$, the machine~$Y$ may simulate~$X$
and can therefore determine arbitrarily many rational approximations~$X(m)$.
However, within finite time, the machine~$Y$ can only acquire finitely many
such approximations. Hence a function such as the signum function, for which
even rough rational approximations of~$\operatorname{sgn}(x)$ require infinite
precision in the input~$x$, do not exist in the \effective topos.

\subsection*{The formal translation rules} The previous examples were picked to
convey an intuitive understanding of what statements in the \effective topos
externally mean, and to showcase a variety of different situations. The
formal translation rules are given in Table~\ref{table:eff}.

\begin{table}
  \begin{tabbing}
    $\Eff \models (\forall n\?\NN\_ \varphi(n))\!\!\!$ \= \kill
    $\Eff \models \varphi$ \> iff there is a Turing machine~$M$ such that~$M
    \realizes \varphi$. \\\\
    $M \realizes \top$ \> is true for any machine~$M$. \\
    $M \realizes \bot$ \> is false for any machine~$M$. \\
    $M \realizes (\varphi \wedge \psi)$ \> iff~$M$ terminates on inputs~$0$
    and~$1$ such that $M(0) \realizes \varphi$ and~$M(1) \realizes \psi$. \\
    $M \realizes (\varphi \vee \psi)$ \> iff~$M$ terminates on inputs~$0$ and~$1$
    such that, \\ \> \qquad if~$M(0) = 0$ then~$M(1) \realizes \varphi$, and if~$M(0) \neq
    0$ then~$M(1) \realizes \psi$. \\
    $M \realizes (\varphi \Rightarrow \psi)$ \> iff for any Turing machine~$E$
    such that~$E \realizes \varphi$, \\ \> \qquad $M$ terminates on input~$E$ such
    that~$M(E) \realizes \psi$. \\
    $M \realizes (\forall n\?\NN\_ \varphi(n))$ \> iff for any natural number~$n_0
    \in \NN$, \\ \> \qquad $M$ terminates on input~$n_0$ such that~$M(n_0)
    \realizes \varphi(n_0)$. \\
    $M \realizes (\exists n\?\NN\_ \varphi(n))$ \> iff~$M$ terminates on
    inputs~$0$ and~$1$ such that $M(1) \realizes \varphi(M(0))$.
  \end{tabbing}

  \caption{\label{table:eff} The (first-order fragment of) the translation
  rules defining the meaning of statements internal to the \effective topos.}
\end{table}
